---
title: "2_known_prep_analysis"
author: "Ajay Prakash Nair"
date: "2024-11-14"
output: html_document
---


```{r setup, include=FALSE}
# Set CRAN mirror
options(repos = c(CRAN = "https://cran.rstudio.com/"))

# Other setup configurations
knitr::opts_chunk$set(echo = TRUE)
```

# Requirements

```{r requirements, results='hide'}
requirements=c("summarytools", "pROC", "glmnetUtils", "dplyr", "car", "effects", "gridExtra", "grid", "MASS","e1071", "mgcv", "caret")

for (req in requirements){
  if (!require(req, character.only = TRUE)){
      install.packages(req)
  }
}
```

# Analysis Description

This analysis focuses on building and evaluating predictive models using logistic regression and Naive Bayes for the dataset. The dataset comprises training and testing data that was splitted in the previous stage. Logistic regression is employed to predict outcomes based on a set of predictor variables. The objective is to assess the predictive performance of the models and evaluate their ability to classify outcomes accurately.

# Loading data

```{r}
train_data <- read.csv("./training/training.csv")
cat("Training data dimensions:", dim(train_data), "\n")

test_data <- read.csv("./testing/test.csv")
cat("Testing data dimensions:", dim(test_data), "\n")
```